---
title: "Build classifier for TB susceptibility"
date: 2016-07-26
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```



## Setup

```{r packages, message=FALSE}
library("caret")
library("dplyr")
library("limma")
library("ggplot2")
library("ggbeeswarm")
library("cowplot")
theme_set(theme_bw(base_size = 14))
theme_update(panel.grid.minor.x = element_blank(),
             panel.grid.minor.y = element_blank(),
             panel.grid.major.x = element_blank(),
             panel.grid.major.y = element_blank())
```

Input limma voom result and annotation.
Note that importing the voom object is just a convenient way to get the TMM-normalized log~2~ cpm data.
The voom weights are not being used.

```{r input-data}
v <- readRDS("../data/results-limma-voom.rds")
anno <- read.delim("../data/experiment-info-filtered.txt",
                   stringsAsFactors = FALSE, row.names = 1)
stopifnot(colnames(v$E) == rownames(anno))
```

Input ash results.

```{r input-ash}
results <- readRDS("../data/results-limma-stats.rds")
```

## Build model to predict susceptibility status

Select those genes that are DE by status or interaction.

```{r}
genes_for_classifer <- v$E[results[["diff_before"]]$svalue < .05 |
                           results[["diff_treat"]]$svalue < .05, ]
# genes_for_classifer <- v$E
# genes_for_classifer <- genes_for_classifer[order(ash_status$svalue)[1:3000], ]
genes_for_classifer <- t(genes_for_classifer)
dim(genes_for_classifer)
genes_for_classifer[1:5, 1:5]
```

Add susceptibility status.

```{r}
genes_for_classifer <- as.data.frame(genes_for_classifer)
genes_for_classifer$status <- anno$status
```

Only include the samples before treatment.

```{r}
genes_for_classifer <- genes_for_classifer[anno$treatment == "none", ]
# stopifnot(nrow(genes_for_classifer) == 25)
```

Fit SVM.

http://stackoverflow.com/questions/20461476/svm-with-cross-validation-in-r-using-caret

http://topepo.github.io/caret/training.html

```{r}
# ctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 10,
#                      savePred = "final", classProbs = TRUE)
ctrl <- trainControl(method = "LOOCV", classProbs = TRUE, savePred = "final")
set.seed(12345)
mod_svm <- train(status ~ ., data = genes_for_classifer,
             method = "svmLinear", trControl = ctrl, metric = "kappa")
mod_svm
head(mod_svm$pred)
```

How well did it work?

```{r}
# methods(class = class(mod_svm))
# names(mod_svm)
# confusionMatrix(mod_svm)
mod_svm$results
```

Use an elastic net implemented in glmnet package.

```{r}
set.seed(12345)
mod_glmnet <- train(status ~ ., data = genes_for_classifer,
                    method = "glmnet", trControl = ctrl, metric = "kappa")
mod_glmnet
head(mod_glmnet$pred)
# confusionMatrix(mod_glmnet)
mod_glmnet$results
```

Visualizing results.

```{r}
d_glmnet <- mod_glmnet$pred %>% mutate(obs_new = ifelse(obs == "tb", "susceptible", "resistant"),
                                       pred_new = ifelse(pred == "tb", "susceptible", "resistant"),
                                       result = paste0("Observed: ", obs_new, "\nPredicted: ", pred_new))
ggplot(d_glmnet, aes(x = result, y = tb)) +
  geom_boxplot() + #geom_point() + geom_beeswarm() +
  labs(x = "Classification", y = "Assigned probability of being susceptible to TB")
```

```{r}
d_glmnet$id <- rownames(genes_for_classifer)[d_glmnet$rowIndex]
d_glmnet <- d_glmnet %>% arrange(contact)
plot(d_glmnet$contact, col = ifelse(d_glmnet$obs == "tb", "red", "black"))
text(x = 1:nrow(d_glmnet) + 0.5, y = d_glmnet$contact,
     labels = substr(d_glmnet$id, 1, 4))
```

Will test on data from Luis's PNAS paper:

```{r}
load("../data/Exp_final_Batch_corrected.Rdata")
```

## Session information

```{r info}
sessionInfo()
```
