---
title: "Compare DE results to eQTL results"
date: 2016-08-25
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

Are DE genes more likely to be an eQTL in this system?

## Setup

```{r packages, message=FALSE}
library("plyr")
library("dplyr")
library("gdata")
library("ggplot2")
library("cowplot")
```

Input DE statistics from limma analysis.

```{r input-limma}
results <- readRDS("../data/results-limma-stats.rds")
results_df <- ldply(results, .id = "contrast")
# Confirm order is maintained before adding back gene names
stopifnot(results_df$t == sapply(results, function(x) x$t))
results_df$gene <- rownames(results[[1]])
head(results_df)
```

Process supplemental data from PNAS paper.
[Barreiro et al., 2012][barreiro2012] infected dendritic cells (DCs) with MTB and identified response eQTLs.
Table S3 contains the eQTL results.

[barreiro2012]: http://www.pnas.org/content/109/4/1204.long

```{r input-eqtl}
eqtl_fname <- "../data/eqtl.txt"
if (file.exists(eqtl_fname)) {
  dc_eqtl <- read.delim(eqtl_fname)
} else {
  dc_eqtl <- read.xls("http://www.pnas.org/content/suppl/2012/01/10/1115761109.DCSupplemental/sd03.xls",
                      skip = 2, nrow = 11954, stringsAsFactors = FALSE)
  write.table(dc_eqtl, file = eqtl_fname, quote = FALSE, sep = "\t")  
}
dc_eqtl[1, ]
```

Classify eQTL type.

```{r classify-eqtl}
dc_eqtl$eqtl_type <- NA
for (i in 1:nrow(dc_eqtl)) {
  if (dc_eqtl$General.eQTL_2f[i] == "YES") {
    dc_eqtl$eqtl_type[i] <- "general"
  } else if (dc_eqtl$NI.Only.eQTL_2f[i] == "YES") {
    dc_eqtl$eqtl_type[i] <- "pre"
  } else if (dc_eqtl$MTB.Only.eQTL_2f[i] == "YES") {
    dc_eqtl$eqtl_type[i] <- "post"
  } else {
    dc_eqtl$eqtl_type[i] <- "non"
  }
}
table(dc_eqtl$eqtl_type)
```

Combine the two data sets.

```{r combine}
d <- merge(results_df, dc_eqtl, by.x = "gene", by.y = "Ensembl.gene.ID")
table(d$eqtl_type[d$contrast == "diff_before"])
```

## Are susceptibility DE genes more likely to have an eQTL?

```{r}
ggplot(d, aes(x = logFC, y = P.value.eQTL.NI.samplesb)) +
  geom_point(alpha = 0.5, aes(color = eqtl_type)) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  facet_grid(eqtl_type ~ contrast)
```

```{r}
ggplot(d[d$eqtl_type != "non" &
         d$contrast %in% c("diff_before", "diff_after"), ],
       aes(x = abs(logFC), y = P.value.eQTL.NI.samplesb)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(eqtl_type ~ contrast) +
  scale_x_log10() +
  scale_y_log10()
```

What are those slopes?

```{r}
x <- d %>%
  filter(eqtl_type != "non") %>% #,
             # contrast %in% c("diff_before", "diff_after")) %>%
  group_by(eqtl_type, contrast) %>%
  do(mod = lm(P.value.eQTL.NI.samplesb ~ abs(logFC), data = .)) %>%
  do(data.frame(eqtl_type = .$eqtl_type,
                contrast = .$contrast,
                slope = coef(.$mod)[2],
                t = summary(.$mod)$coefficients[2, 3],
                p = summary(.$mod)$coefficients[2, 4],
                r2 = summary(.$mod)$r.squared))
ggplot(x, aes(x = eqtl_type, y = slope)) + geom_point() + facet_wrap(~contrast)
ggplot(x, aes(x = eqtl_type, y = abs(t))) + geom_point() + facet_wrap(~contrast)
ggplot(x, aes(x = eqtl_type, y = p)) + geom_point() + facet_wrap(~contrast)
ggplot(x, aes(x = eqtl_type, y = r2)) + geom_point() + facet_wrap(~contrast)
```


```{r}
ggplot(d[d$eqtl_type %in% c("pre", "post", "general") &
         d$contrast %in% c("diff_before", "diff_after"), ],
       aes(x = eqtl_type, y = abs(logFC))) +
  geom_boxplot() +
  facet_wrap(~contrast) +
  scale_y_log10()
```

```{r}
d %>%
  filter(eqtl_type %in% c("pre", "post", "general", "non"),
             contrast %in% c("diff_before", "diff_after")) %>%
  group_by(eqtl_type, contrast) %>%
  summarize(lfc = median(abs(logFC)),
            exp_lev = mean(AveExpr),
            lfc_sem = sd(abs(logFC)) / n()) %>%
  ggplot(., aes(x = eqtl_type, y = lfc)) +
  geom_point() +
  facet_wrap(~contrast) +
  geom_linerange(aes(ymin = lfc - lfc_sem,
                    ymax = lfc + lfc_sem)) +
  geom_text(aes(label = lfc))

```

```{r}
d %>%
  group_by(eqtl_type, contrast) %>%
  summarize(lfc = median(abs(logFC)),
            exp_lev = mean(AveExpr),
            lfc_sem = sd(abs(logFC)) / n()) %>%
  ggplot(., aes(x = eqtl_type, y = lfc)) +
  geom_point() +
  facet_wrap(~contrast, scales = "free_y") +
  geom_linerange(aes(ymin = lfc - lfc_sem,
                    ymax = lfc + lfc_sem))
```

## Overlap of DE genes and genes with eQTLs

Is there a significant overlap in genes that are differentially expressed by susceptibility status and genes that have a response-eQTL?

```{r overlap}
d$de <- ifelse(d$P.Value < 0.05, "de", "not")
table(d$de, d$contrast)
diff_before_de <- d$de[d$contrast == "diff_before" & d$eqtl_type %in% c("pre", "post")]
diff_after_de <- d$de[d$contrast == "diff_after" & d$eqtl_type %in% c("pre", "post")]

factor_eqtl <- factor(d$eqtl_type[d$eqtl_type %in% c("pre", "post") &
                                  d$contrast == "diff_before"],
                      levels = c("pre", "post"))
table(factor_eqtl)
table(factor_eqtl, diff_before_de)
chisq.test(x = factor_eqtl, y = diff_before_de)
table(factor_eqtl, diff_after_de)
chisq.test(x = factor_eqtl, y = diff_after_de)
```

Genes that are DE by susceptibility status and a response-eQTL in the concordant condition.

```{r genes-concordant}
d %>%
  filter(eqtl_type == "pre", contrast == "diff_before", de == "de") %>%
  select(gene) %>% arrange(gene)
d %>%
  filter(eqtl_type == "post", contrast == "diff_after", de == "de") %>%
  select(gene) %>% arrange(gene)
```

Genes that are DE by susceptibility status and a response-eQTL in the disconcordant condition.

```{r genes-discordant}
d %>%
  filter(eqtl_type == "pre", contrast == "diff_after", de == "de") %>%
  select(gene) %>% arrange(gene)
d %>%
  filter(eqtl_type == "post", contrast == "diff_before", de == "de") %>%
  select(gene) %>% arrange(gene)
```

## Session information

```{r info}
sessionInfo()
```
