---
title: "Is there an increase in the variance in the infected state?"
date: 2017-01-04
output: html_document
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

## Setup

```{r packages, message=FALSE}
library("limma")
library("dplyr")
library("tidyr")
library("ggplot2")
library("cowplot")
```

Load limma results produced by file `../code/main-limma.R`.
The log~2~ cpm are in the Elist object created by the voom function (`v2`).

```{r load-limma-results}
v2 <- readRDS("../data/results-limma-voom.rds")
fit2 <- readRDS("../data/results-limma-fit.rds")
results <- readRDS("../data/results-limma-stats.rds")
d <- v2$E
```

## Calculate the variance

For each gene, calculate the mean, variance, and coefficient of variation (CV) in the non-infected and infected state.

```{r}
n_genes <- nrow(d)
m_mean <- matrix(nrow = n_genes, ncol = 2,
                 dimnames = list(rownames(d), c("noninf", "infect")))
m_var <- matrix(nrow = n_genes, ncol = 2,
                dimnames = list(rownames(d), c("noninf", "infect")))
m_mean_r <- matrix(nrow = n_genes, ncol = 2,
                 dimnames = list(rownames(d), c("noninf", "infect")))
m_var_r <- matrix(nrow = n_genes, ncol = 2,
                dimnames = list(rownames(d), c("noninf", "infect")))
m_mean_s <- matrix(nrow = n_genes, ncol = 2,
                 dimnames = list(rownames(d), c("noninf", "infect")))
m_var_s <- matrix(nrow = n_genes, ncol = 2,
                dimnames = list(rownames(d), c("noninf", "infect")))
ftest_p <- numeric(length = n_genes)
names(ftest_p) <- rownames(d)
ftest_stat <- numeric(length = n_genes)
names(ftest_stat) <- rownames(d)
ftest_p_r <- numeric(length = n_genes)
names(ftest_p_r) <- rownames(d)
ftest_stat_r <- numeric(length = n_genes)
names(ftest_stat_r) <- rownames(d)
ftest_p_s <- numeric(length = n_genes)
names(ftest_p_s) <- rownames(d)
ftest_stat_s <- numeric(length = n_genes)
names(ftest_stat_s) <- rownames(d)

for (i in 1:n_genes) {
  d_noninf <- d[i, grepl("noninf", colnames(d))]
  d_infect <- d[i, grepl("infect", colnames(d))]
  d_noninf_r <- d_noninf[grepl("resist", names(d_noninf))]
  d_infect_r <- d_infect[grepl("resist", names(d_infect))]
  d_noninf_s <- d_noninf[grepl("suscep", names(d_noninf))]
  d_infect_s <- d_infect[grepl("suscep", names(d_infect))]

  m_mean[i, "noninf"] <- mean(d_noninf)
  m_mean[i, "infect"] <- mean(d_infect)
  m_var[i, "noninf"] <- var(d_noninf)
  m_var[i, "infect"] <- var(d_infect)
  m_mean_r[i, "noninf"] <- mean(d_noninf_r)
  m_mean_r[i, "infect"] <- mean(d_infect_r)
  m_var_r[i, "noninf"] <- var(d_noninf_r)
  m_var_r[i, "infect"] <- var(d_infect_r)
  m_mean_s[i, "noninf"] <- mean(d_noninf_s)
  m_mean_s[i, "infect"] <- mean(d_infect_s)
  m_var_s[i, "noninf"] <- var(d_noninf_s)
  m_var_s[i, "infect"] <- var(d_infect_s)

  ftest <- var.test(d_infect, d_noninf)
  ftest_p[i] <- ftest$p.value
  ftest_stat[i] <- ftest$statistic
  stopifnot(ftest_stat[i] == m_var[i, "infect"] / m_var[i, "noninf"])

  ftest_r <- var.test(d_infect_r, d_noninf_r)
  ftest_p_r[i] <- ftest_r$p.value
  ftest_stat_r[i] <- ftest_r$statistic
  stopifnot(ftest_stat_r[i] == m_var_r[i, "infect"] / m_var_r[i, "noninf"])

  ftest_s <- var.test(d_infect_s, d_noninf_s)
  ftest_p_s[i] <- ftest_s$p.value
  ftest_stat_s[i] <- ftest_s$statistic
  stopifnot(ftest_stat_s[i] == m_var_s[i, "infect"] / m_var_s[i, "noninf"])
}
m_cv <- m_var / m_mean
```

Organize ftest results.

```{r}
df_ftest <- data.frame(gene = rownames(d), ftest_p, ftest_stat)
stopifnot(rownames(d) == rownames(results[["status_ni"]]))
df_ftest$is_de <- results[["status_ni"]]$qvalue < 0.1
stopifnot(sum(df_ftest$is_de) == 645)
```

Organize long df for ggplot of histograms.

```{r}
df_hist <- data.frame(gene = names(ftest_stat),
                all = ftest_stat,
                resist = ftest_stat_r,
                suscep = ftest_stat_s)
stopifnot(rownames(df_hist) == df_hist$gene)
# Convert to long form
df_hist <- gather(df_hist, key = "individuals", value = "var_ratio", -gene)
# Add an indicator if the gene is DE for status_ni (difference in status in
# non-infected state)
df_hist$is_de <- ifelse(results[["status_ni"]]$qvalue < 0.1, TRUE, FALSE)
stopifnot(sum(df_hist$is_de) == 645 * 3)
table(df_hist$individuals, df_hist$is_de)
head(df_hist)
```

Improve factor labels.

```{r}
df_hist$individuals <- factor(df_hist$individuals,
                              levels = c("all", "resist", "suscep"),
                              labels = c(sprintf("All individuals (n = %d)",
                                                 ncol(d) / 2),
                                         sprintf("Resistant individuals (n = %d)",
                                                 sum(grepl("resist", colnames(d))) / 2),
                                         sprintf("Susceptible individuals (n = %d)",
                                                 sum(grepl("suscep", colnames(d))) / 2)))
```

```{r}
df_hist$is_de <- factor(df_hist$is_de,
                        levels = c(FALSE, TRUE),
                        labels = c(sprintf("All genes (n = %d)",
                                           nrow(d)),
                                   sprintf("DE genes in non-infected state (n = %d)",
                                           sum(results[["status_ni"]]$qvalue < 0.1))))
```


## Count

How many genes had higher variance in the infected state?

```{r}
df_ftest %>% filter(ftest_stat > 1) %>% nrow()
# significant
df_ftest %>% filter(ftest_stat > 1, ftest_p < 0.05) %>% nrow()
# of the status_ni DE genes
df_ftest %>% filter(ftest_stat > 1, is_de) %>% nrow()
# of the status_ni DE genes and significant
df_ftest %>% filter(ftest_stat > 1, ftest_p < 0.05, is_de) %>% nrow()
```

How many genes had higher variance in the non-infected state?

```{r}
df_ftest %>% filter(ftest_stat < 1) %>% nrow()
# significant
df_ftest %>% filter(ftest_stat < 1, ftest_p < 0.05) %>% nrow()
# of the status_ni DE genes
df_ftest %>% filter(ftest_stat < 1, is_de) %>% nrow()
# of the status_ni DE genes and significant
df_ftest %>% filter(ftest_stat < 1, ftest_p < 0.05, is_de) %>% nrow()
```

## Visualize

```{r supp-figure}
ggplot(df_hist, aes(x = log2(var_ratio))) +
  geom_histogram(binwidth = 0.25) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  facet_grid(is_de ~ individuals, scales = "free_y") +
  labs(x = expression(log[2] * "(" * frac("Variance in the infected state", "Variance in the non-infected state") * ")"),
       y = "Number of genes")
```



```{r compare-mean}
ggplot(as.data.frame(m_mean), aes(x = noninf, y = infect)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red")
```


```{r compare-var}
ggplot(as.data.frame(m_var), aes(x = noninf, y = infect)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  scale_x_sqrt() + scale_y_sqrt()
```

```{r boxplot-var}
boxplot(sqrt(m_var))
```

```{r compare-cv}
ggplot(as.data.frame(m_cv), aes(x = noninf, y = infect)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  scale_x_sqrt() + scale_y_sqrt()
```

```{r}
hist(ftest_p)
hist(ftest_stat)
hist(log2(ftest_stat))
plot(ftest, results[["status_ni"]]$P.Value)
plot(ftest, results[["status_ii"]]$P.Value)
```

```{r}

ggplot(df_ftest,
       aes(x = log2(ftest_stat), y = -log10(ftest_p),
           color = is_de, shape = is_de)) +
  geom_point() +
  scale_shape_manual(values = c(1, 16), drop = FALSE) +
  scale_color_manual(values = c("gray75", "darkred"), drop = FALSE) +
  facet_wrap(~is_de)
```

## Session information

```{r info}
sessionInfo()
```
