---
title: "Investigating the interaction term"
date: 2017-01-04
output: html_document
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

I ignore the interaction term for most analyses because it is uninteresting.
Evidence below.
Mainly, because there are DE genes between resistant and susceptible individuals in the non-infected state but not the infected state, the interaction term is partially redundant with the contrast of susceptibility status in the non-infected state (Pearson's r ~= 0.27 between their p-values).

## Setup

```{r packages, message=FALSE}
library("limma")
library("plyr")
library("dplyr")
library("tidyr")
library("ggplot2")
library("cowplot")
library("gplots")
```

Load limma results produced by file `../code/main-limma.R`.

```{r load-limma-results}
v2 <- readRDS("../data/results-limma-voom.rds")
fit2 <- readRDS("../data/results-limma-fit.rds")
results <- readRDS("../data/results-limma-stats.rds")
```

## Overlap of significant genes

```{r de-genes}
ldply(results, function(x) sum(x$qvalue < .1), .id = "test")
de_list <- llply(results, function(x) rownames(x)[x$qvalue < .1])
lapply(de_list, length)
```

There is only 1 gene with a qvalue less than 10% for the interaction term.
It is also significant for every test except for status_ni (resistant vs. susceptible in the infected state).

```{r}
(interaction_sub <- subset(results[["interact"]], qvalue < .1))
interaction_gene <- rownames(interaction_sub)
stopifnot(length(interaction_gene) == 1)
(interaction_gene_results <- ldply(results, function(x) x[interaction_gene, ], .id = "test"))
interaction_gene_results$test[interaction_gene_results$qvalue < 0.1]
```

[ENSG00000086061 at Ensembl](http://useast.ensembl.org/Homo_sapiens/Gene/Summary?db=core;g=ENSG00000086061;r=9:33025211-33039907;t=ENST00000330899)

## Visualization

Visualizing the one gene that is significantly DE for the interaction term.

```{r plot-gene}
plot_gene <- function(v, g) {
  # v - An EList object containing log2 counts per million
  # g - character vector of a single gene
  stopifnot(class(v) == "EList",
            is.character(g), length(g) == 1)
  library("tidyr")
  single_gene <- v$E[g, ]
  single_gene_long <- gather_(as.data.frame(single_gene),
                              gather_cols = colnames(single_gene))
  # For some reason, the argument value_col wouldn't work
  colnames(single_gene_long) <- "log2cpm"
  single_gene_long$sample <- rownames(single_gene_long)
  single_gene_long <- separate_(single_gene_long, col = "sample", sep = "-",
                                into = c("individual", "status", "treatment"))
  single_gene_long$status <- factor(single_gene_long$status, levels = c("resist", "suscep"))
  single_gene_long$treatment <- factor(single_gene_long$treatment, levels = c("noninf", "infect"))
  ggplot(single_gene_long, aes(x = treatment, y = log2cpm, fill = status)) +
    geom_boxplot() +
    labs(title = g, x = "Treatment", y = expression("Expression level (" * log[2] * " cpm)"))
}
```

As expected from the DE results, the difference in the infected state is not very impressive.
Likely the interaction term is significant mainly because the direction of the fold difference between susceptible and resistant individuals is in the opposite direction for the non-infected and infected states.

```{r plot-gene-interaction}
plot_gene(v2, interaction_gene)
```

## Histogram of p-values

```{r}
# http://stackoverflow.com/a/3562030/2483477
l_ply(names(results), function(x) hist(results[[x]]$P.Value, main = x))
```

## Pairwise relationship of p-values

```{r}
pval <- do.call(cbind, lapply(results, function(x) x$P.Value))
(pval_cor <- cor(pval))
heatmap.2(pval_cor, trace = "none", margins = c(9, 9))
pairs(-log10(pval))
```

## Relationship between logFC of contrasts and significance of interaction term

status_ni

```{r}
cor(abs(results[["status_ni"]]$logFC), -log10(results[["interact"]]$P.Value))
plot(abs(results[["status_ni"]]$logFC), -log10(results[["interact"]]$P.Value))
abline(0, 1, col = "red")
```

status_ii

```{r}
cor(abs(results[["status_ii"]]$logFC), -log10(results[["interact"]]$P.Value))
plot(abs(results[["status_ii"]]$logFC), -log10(results[["interact"]]$P.Value))
abline(0, 1, col = "red")
```

treat_resist

```{r}
cor(abs(results[["treat_resist"]]$logFC), -log10(results[["interact"]]$P.Value))
plot(abs(results[["treat_resist"]]$logFC), -log10(results[["interact"]]$P.Value))
abline(0, 1, col = "red")
```

treat_suscep

```{r}
cor(abs(results[["treat_suscep"]]$logFC), -log10(results[["interact"]]$P.Value))
plot(abs(results[["treat_suscep"]]$logFC), -log10(results[["interact"]]$P.Value))
abline(0, 1, col = "red")
```

interact (best-case scenario)

```{r}
cor(abs(results[["interact"]]$logFC), -log10(results[["interact"]]$P.Value))
plot(abs(results[["interact"]]$logFC), -log10(results[["interact"]]$P.Value))
abline(0, 1, col = "red")
```

## Session information

```{r info}
sessionInfo()
```
