---
title: "Check for discordant lanes"
date: 2016-08-03
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

## Setup

```{r packages, message=FALSE}
library("data.table")
library("dplyr")
library("edgeR")
library("gplots")
library("ggplot2")
theme_set(theme_bw(base_size = 12))
theme_update(panel.grid.minor.x = element_blank(),
             panel.grid.minor.y = element_blank(),
             panel.grid.major.x = element_blank(),
             panel.grid.major.y = element_blank())
```

Input raw counts file.

```{r input-counts}
data_raw <- fread("../data/subread-counts-per-lane.txt", data.table = FALSE)
data_raw[1:5, 1:7]
```

Split annotation and counts.

```{r}
anno <- data_raw %>% select(individual:lane)
counts <- data_raw %>% select(starts_with("ENSG")) %>% t()
anno$id <- paste(anno$individual, anno$treatment, anno$flow_cell, sep = ".")
colnames(counts) <- anno$id
```

Standardize by library size.

```{r}
counts_cpm <- cpm(counts, log = TRUE)
```

## First pass

Perform PCA.

```{r}
pca <- prcomp(t(counts_cpm), scale. = TRUE)
variances <- pca$sdev^2
explained <- variances / sum(variances)
plot(pca, main = "Variance per PC")
pca_data <- cbind(anno, pca$x)
```

```{r pc1-pc2}
ggplot(pca_data, aes(x = PC1, y = PC2, color = treatment)) +
  geom_text(aes(label = individual)) +
  labs(x = sprintf("PC%d (%.2f%%)", 1, round(explained[1] * 100, 2)),
       y = sprintf("PC%d (%.2f%%)", 2, round(explained[2] * 100, 2)))
```

```{r pc3-pc4}
ggplot(pca_data, aes(x = PC3, y = PC4, color = treatment)) +
  geom_text(aes(label = individual)) +
  labs(x = sprintf("PC%d (%.2f%%)", 3, round(explained[3] * 100, 2)),
       y = sprintf("PC%d (%.2f%%)", 4, round(explained[4] * 100, 2)))
```

Perform hierachical clustering.

```{r fig.width=8, fig.height=8}
counts_cor <- cor(counts_cpm)
clust <- heatmap.2(counts_cor, trace = "none", margins = c(7, 7))
```

From the PCA and hierarchical clustering, there are two main outliers: 

```{r}
(outliers <- colnames(counts_cor)[clust$colInd[1:2]])
```

This is strange.
Why would two lanes of sequencing from this sample be different from the third?
This is unexpected because they are from the same library preparation.
I suspected this was due to sequencing depth.
This is because 03-contact-infected was one of the three samples that was over-sequenced in the first round.
After adjusting the amount of 03-contact-infected in the master mix, very little was sequenced in the subsequent two rounds.

```{r}
depth <- colSums(counts)
depth <- sort(depth)
summary(depth)
head(depth)
tail(depth)
```

This was clearly the reason.
It's first lane was one of the most highly sequenced, and its two subsequent lanes were the two least sequenced.
This is concordant with the sequencing summary statistics provided by the core.

## Second pass

Becuase they are outliers due to depth, and not a problem with the library itself, I think it makes sense to keep them in the final analysis.
However, in order to better investigate differences across lanes, I remove them here.

```{r}
counts_cpm <- counts_cpm[, !(colnames(counts_cpm) %in% outliers)]
anno <- anno[!(anno$id %in% outliers), ]
```

Perform PCA.

```{r}
pca <- prcomp(t(counts_cpm), scale. = TRUE)
variances <- pca$sdev^2
explained <- variances / sum(variances)
plot(pca, main = "Variance per PC")
pca_data <- cbind(anno, pca$x)
```

```{r pc1-pc2-filtered}
ggplot(pca_data, aes(x = PC1, y = PC2, color = treatment)) +
  geom_text(aes(label = individual)) +
  labs(x = sprintf("PC%d (%.2f%%)", 1, round(explained[1] * 100, 2)),
       y = sprintf("PC%d (%.2f%%)", 2, round(explained[2] * 100, 2)))
```

```{r pc3-pc4-filtered}
ggplot(pca_data, aes(x = PC3, y = PC4, color = treatment)) +
  geom_text(aes(label = individual)) +
  labs(x = sprintf("PC%d (%.2f%%)", 3, round(explained[3] * 100, 2)),
       y = sprintf("PC%d (%.2f%%)", 4, round(explained[4] * 100, 2)))
```

```{r pc5-pc6-filtered}
ggplot(pca_data, aes(x = PC5, y = PC6, color = treatment)) +
  geom_text(aes(label = individual)) +
  labs(x = sprintf("PC%d (%.2f%%)", 5, round(explained[5] * 100, 2)),
       y = sprintf("PC%d (%.2f%%)", 6, round(explained[6] * 100, 2)))
```

Perform hierachical clustering.

```{r fig.width=8, fig.height=8}
counts_cor <- cor(counts_cpm)
clust <- heatmap.2(counts_cor, trace = "none", margins = c(7, 7))
```

There are three main clusters.
The second is clearly the controls ("none") and the third is clearly the infected samples.
The first cluster are the misfits.

```{r}
(misfits <- colnames(counts_cor)[clust$rowInd[1:26]])
```

The first 3 sets are not surprising given the PCA results.
All 3 replicates of c06.infected, c06.none, and c02.none are in this cluster,
consistent with them being outliers in the PCA.
The others in this cluster are harder to interpret.
The only other one with all 3 replicates represented is c04.none.

```{r}
table(unlist(strsplit(misfits, ".H")))
```

Maybe it is a sequencing depth issue again.

```{r}
misfits_unknown <- misfits[!(substr(misfits, 1, 3) %in% c("c06", "c02"))]
misfit_factor <- ifelse(names(depth) %in% misfits_unknown, TRUE, FALSE)
boxplot(log(depth) ~ misfit_factor)
```

It appears that sequencing depth is contributing to the issue.

## Session information

```{r info}
sessionInfo()
```
