---
title: "Setup"
date: 2016-07-07
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

Prepare data files.

## Setup

```{r packages, message=FALSE}
library("data.table")
library("dplyr")
library("edgeR")
```

Input raw counts file.

```{r input-counts}
data_raw <- fread("../data/subread-counts-per-sample.txt", data.table = FALSE)
data_raw[1:5, 1:5]
```

Input experimental information file.

```{r experimental-info}
info <- read.delim("../data/experiment-info.txt", stringsAsFactors = FALSE)
head(info)
```

Sort raw data file so that it is in the same order as the information file.

```{r sort-data-raw}
data_raw <- data_raw %>% arrange(desc(status), individual, desc(treatment))
stopifnot(data_raw$inindividual == info$individual,
          data_raw$status == info$status,
          data_raw$treatment == info$treatment)
```  

Create a sample-x-gene counts matrix.

```{r transpose-counts}
counts_raw <- data_raw %>% select(-(individual:treatment)) %>% t()
colnames(counts_raw) <- info$id
stopifnot(ncol(counts_raw) == 50)
```

## Remove lowly expressed genes

```{r median-density}
cpm_raw <- cpm(counts_raw, log = TRUE)
cpm_raw_median <- apply(cpm_raw, 1, median)
cpm_raw_median_density <- density(cpm_raw_median)
cpm_raw_density <- apply(cpm_raw, 2, density)
plot(cpm_raw_density[[1]], lty = "dotted", col = gray(0.05, alpha = 0.5),
     main = "Gene expression distribution",
     xlab = "log2 cpm", sub = "red = median; blue = cutoff")
suppress_output <- lapply(cpm_raw_density, lines, lty = "dotted",
                          col = gray(0.05, alpha = 0.5))
lines(cpm_raw_median_density, col = "red", lwd = 3)
cutoff <- 0
abline(v = cutoff, col = "blue")
```

Keeping the `r sum(cpm_raw_median > cutoff)` genes with a median log~2~ cpm greater than `r cutoff`.

```{r filter-genes}
counts <- as.data.frame(counts_raw) %>% filter(cpm_raw_median > cutoff)
rownames(counts) <- rownames(counts_raw)[cpm_raw_median > cutoff]
```

Save the count matrix.

```{r}
write.table(counts, "../data/counts.txt", quote = FALSE, sep = "\t",
            col.names = NA)
```

## Session information

```{r info}
sessionInfo()
```
