---
title: "Test ash models"
date: 2016-07-28
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

This analysis compares two different models for ash: exchangeable effects (EE) and exchangeable standardized effects (ET) models.
For more details on the assumptions behind the models, see this [analysis by Matthew Stephens][stephens].

In the analysis below, the ET model has a larger log likelihood compared to the EE model.

[stephens]: http://stephenslab.github.io/ash/analysis/CompareEEvsETmodel.html

## Setup

```{r packages, message=FALSE}
library("limma")
library("ashr")
```

Input the results of the model fit by limma.

```{r input-counts}
fit <- readRDS("../data/results-limma-fit.rds")
```

The linear model was the following:

$$ Y\ \sim\beta_0+X_{treat}\beta_{treat}+X_{status}\beta_{status}+X_{treat,status}\beta_{treat,status}+I+\epsilon $$

where $Y$ is the expression level of a gene,
$\beta_{treat}$ is the fixed effect of treatment with MTB,
$\beta_{status}$ is the fixed effect of susceptibility status,
$\beta_{treat,status}$ is the fixed effect of interaction between treatment and susceptibilty status,
and $I$ is the random effect of individual (implemented via `duplicateCorrelation`).

The following functions extracts the estimated effect sizes, standard errors of the effect size, and the degrees of freedom from the limma object.

```{r extraction-functions}
extract_betahat_se_df <- function(model, coef) {
  # model - MArrayLM object fit by limma::eBayes
  # coef - coefficient in linear model (column name in design matrix)
  # 
  # Returns list with the estimated betas (betahat), standard errors of beta
  # (sebetahat), and degrees of freedom (df)
  stopifnot(class(model) == "MArrayLM",
            coef %in% colnames(model$design))
  betahat <- model$coefficients[, coef]
  sebetahat <- model$stdev.unscaled[, coef] * sqrt(model$s2.post)
  df <- model$df.total
  return(list(betahat = betahat, sebetahat = sebetahat, df = df))
}
```

## Compare EE versus ET models

### Treatment effect

```{r ash-treatment, results='hide'}
results_treatment <- extract_betahat_se_df(fit, "infected")
ash_treatment_ee <- ash(betahat = results_treatment$betahat,
                        sebetahat = results_treatment$sebetahat,
                        df = results_treatment$df,
                        method = "fdr", model = "EE")
ash_treatment_et <- ash(betahat = results_treatment$betahat,
                        sebetahat = results_treatment$sebetahat,
                        df = results_treatment$df,
                        method = "fdr", model = "ET")
```

```{r ash-treatment-results}
sum(ash_treatment_ee$svalue < .05) / nrow(fit$coefficients)
sum(ash_treatment_et$svalue < .05) / nrow(fit$coefficients)
ash_treatment_ee$loglik - ash_treatment_et$loglik
```

### Susceptibility status effect

```{r ash-status, results='hide'}
results_status <- extract_betahat_se_df(fit, "tb")
ash_status_ee <- ash(betahat = results_status$betahat,
                     sebetahat = results_status$sebetahat,
                     df = results_status$df,
                     method = "fdr", model = "EE")
ash_status_et <- ash(betahat = results_status$betahat,
                     sebetahat = results_status$sebetahat,
                     df = results_status$df,
                     method = "fdr", model = "ET")
```

```{r ash-status-results}
sum(ash_status_ee$svalue < .05)
sum(ash_status_et$svalue < .05)
ash_status_ee$loglik - ash_status_et$loglik
```

### Interaction effect

```{r ash-interaction, results='hide'}
results_interaction <- extract_betahat_se_df(fit, "infected.tb")
ash_interaction_ee <- ash(betahat = results_interaction$betahat,
                          sebetahat = results_interaction$sebetahat,
                          df = results_interaction$df,
                          method = "fdr", model = "EE")
ash_interaction_et <- ash(betahat = results_interaction$betahat,
                          sebetahat = results_interaction$sebetahat,
                          df = results_interaction$df,
                          method = "fdr", model = "ET")
```

```{r ash-interaction-results}
sum(ash_interaction_ee$svalue < .05)
sum(ash_interaction_et$svalue < .05)
ash_interaction_ee$loglik - ash_interaction_et$loglik
```

## Session information

```{r info}
sessionInfo()
```
