---
title: "Analysis of Thuong et al., 2008"
date: 2016-05-27
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

http://www.ncbi.nlm.nih.gov/pubmed/19057661

http://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1000229#s2

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE11199

http://www.bioconductor.org/packages/release/bioc/html/affy.html

http://bioconductor.org/packages/release/bioc/vignettes/GEOquery/inst/doc/GEOquery.html

http://www.bioconductor.org/packages/release/data/annotation/html/hgu133plus2.db.html

## Setup

```{r packages, message=FALSE, warning=FALSE}
library("GEOquery")
library("affy")
library("hgu133plus2.db")
library("limma")
library("tidyr")
library("biomaRt")
```

## Analyze starting from processed data


```{r download}
geo_fname <- "../data/GSE11199.rds"
if (!file.exists(geo_fname)) {
  gse <- getGEO(GEO = "GSE11199")
#   class(gse)
#   length(gse)
  gse <- gse[[1]]
  saveRDS(gse, geo_fname)
} else {
  gse <- readRDS(geo_fname)
}
class(gse)
pData(gse)[1, ]
```

```{r}
x <- exprs(gse)
anno <- pData(gse)[, 1]
anno <- as.data.frame(anno)
anno <- separate(anno, col = anno, into = c("individual", "treatment"))
anno$treatment <- factor(anno$treatment, levels = c("unstim", "stim"),
                         labels = c("none", "infected"))
anno$status <- substr(anno$individual, 1, 1)
anno$status <- factor(anno$status, levels = c("L", "P", "T"),
                      labels = c("latent", "pulmonary", "meningeal"))
table(anno$treatment, anno$status)
```

I tried using the hgu133plus2.db package, but this wasn't very useful.

```{r hgu133plus2}
# columns(hgu133plus2.db)
probe_mappings <- select(hgu133plus2.db, keys = rownames(x),
                         columns = c("ENSEMBL", "SYMBOL"))
head(probe_mappings)
probe_mappings[probe_mappings$SYMBOL == "DDR1" &
               !is.na(probe_mappings$SYMBOL), ]
```

The multipe mappings is because it is reporting lots of Ensembl gene IDs per each probe.
However, when I check these on Ensembl, many are labelled "Not part of the primary assembly".
Thus they should be able to be filtered out by their chromsome name.
If I have to use biomaRt to obtain the chromosome name, I may as well just use it to obtain the probe to gene mappings as well.

```{r probe2id}
ensembl <- useMart(host = "dec2015.archive.ensembl.org",
                   biomart = "ENSEMBL_MART_ENSEMBL",
                   dataset = "hsapiens_gene_ensembl")
probe2id <- getBM(attributes = c("affy_hg_u133_plus_2", "ensembl_gene_id",
                                 "chromosome_name", "status",
                                 "external_gene_name", "gene_biotype"),
                   filters = "affy_hg_u133_plus_2",
                   values = as.character(fData(gse)$ID),
                   mart = ensembl)
head(probe2id)
```


```{r}
probes <- probe2id[probe2id$chromosome_name %in% c(1:22, "X", "Y", "MT") &
                   probe2id$gene_biotype == "protein_coding", ]
nrow(probes)
length(unique(probes$affy_hg_u133_plus_2))
```

Keeping only those probes that unambiguously tag one protein-coding gene.

```{r}
genes_per_probe <- table(probes$affy_hg_u133_plus_2)
good_probes <- names(genes_per_probe)[genes_per_probe == 1]
```

Some genes are tagged by more than one probe.
I'll need to average them somehow to get one value per gene.
The analysis below continues with the probe IDs.

Following limma manual sections 9.3, 9.5.2, and 9.7.

```{r limma}
f <- paste(anno$status, anno$treatment, sep = ".")
f <- factor(f)
design <- model.matrix(~0 + f)
colnames(design) <- levels(f)
colSums(design)
corfit <- duplicateCorrelation(x, design, block = anno$individual)
corfit$consensus
fit <- lmFit(x, design, block = anno$individual, correlation = corfit$consensus)
contrast_matrix <- makeContrasts(
  latent = latent.infected - latent.none,
  pulmonary = pulmonary.infected - pulmonary.none,
  # meningeal = meningeal.infected - meningeal.none,
  interaction = (pulmonary.infected - pulmonary.none) -
                (latent.infected - latent.none),
  levels = design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
results <- decideTests(fit2)
summary(results)
vennDiagram(results)
```

## Analyze starting from raw data

The only information they provided on GEO was "Raw CEL intensity data were RMA normalized using R/Bioconductor".
Thus I just used the recommendation from the affy package example for the other steps.

```{r process-from-raw}
process_raw_fname <- "../data/thuong2008.rds"
if (file.exists(process_raw_fname)) {
  eset <- readRDS(process_raw_fname)
} else {
  # Download
  getGEOSuppFiles(GEO = "GSE11199")
  untar("GSE11199/GSE11199_RAW.tar", exdir = "GSE11199")
  # Import
  raw <- ReadAffy(filenames = Sys.glob("GSE11199/*CEL.gz"), compress = TRUE)
  # Pre-process
  eset <- expresso(raw,
                   bgcorrect.method = "rma",
                   normalize.method = "quantiles",
                   pmcorrect.method = "pmonly",
                   summary.method = "medianpolish")
  saveRDS(eset, process_raw_fname)
}
```

## Compare the two approaches

My pre-processing obtains results very similar to theirs.

```{r comparison}
dim(x)
y <- exprs(eset)
x[1:5, 1:5]
y[1:5, 1:5]
plot(x[, 1], y[, 1])
cor(x[, 1], y[, 1])
z <- cor(cbind(x, y))
dim(z)
min(z)
median(z)
```

## Session information

```{r info}
sessionInfo()
```
