---
title: "Analysis of Thuong et al., 2008"
date: 2016-05-27
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

http://www.ncbi.nlm.nih.gov/pubmed/19057661

http://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1000229#s2

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE11199

http://www.bioconductor.org/packages/release/bioc/html/affy.html

http://bioconductor.org/packages/release/bioc/vignettes/GEOquery/inst/doc/GEOquery.html

http://www.bioconductor.org/packages/release/data/annotation/html/hgu133plus2.db.html

## Setup

```{r packages, message=FALSE, warning=FALSE}
library("GEOquery")
library("affy")
library("tidyr")
library("limma")
library("biomaRt")
library("dplyr")
library("gplots")
```

## Download processed data

```{r download}
geo_fname <- "../data/GSE11199.rds"
if (!file.exists(geo_fname)) {
  gse <- getGEO(GEO = "GSE11199")
#   class(gse)
#   length(gse)
  gse <- gse[[1]]
  saveRDS(gse, geo_fname)
} else {
  gse <- readRDS(geo_fname)
}
class(gse)
x <- exprs(gse)
```

## Download and pre-process the raw data

The only information they provided on GEO was "Raw CEL intensity data were RMA normalized using R/Bioconductor".
Thus I just used the recommendation from the affy package example for the other steps.

```{r process-from-raw}
process_raw_fname <- "../data/thuong2008.rds"
if (file.exists(process_raw_fname)) {
  eset <- readRDS(process_raw_fname)
} else {
  # Download
  getGEOSuppFiles(GEO = "GSE11199")
  untar("GSE11199/GSE11199_RAW.tar", exdir = "GSE11199")
  # Import
  raw <- ReadAffy(filenames = Sys.glob("GSE11199/*CEL.gz"), compress = TRUE)
  # Pre-process
  eset <- expresso(raw,
                   bgcorrect.method = "rma",
                   normalize.method = "quantiles",
                   pmcorrect.method = "pmonly",
                   summary.method = "medianpolish")
  saveRDS(eset, process_raw_fname)
}
```

## The processed data is similar to a standard workflow

Their processed data is very similar to the results of pre-processing the raw data with the recommendations from the affy package, so I'll use their pre-processed data since that is more convenient.

```{r comparison}
dim(x)
y <- exprs(eset)
x[1:5, 1:5]
y[1:5, 1:5]
plot(x[, 1], y[, 1])
cor(x[, 1], y[, 1])
z <- cor(cbind(x, y))
dim(z)
min(z)
median(z)
```

## Prepare sample annotation

```{r sample-annotation}
anno <- pData(gse)[, 1]
anno <- as.data.frame(anno)
anno <- separate(anno, col = anno, into = c("individual", "treatment"))
anno$treatment <- factor(anno$treatment, levels = c("unstim", "stim"),
                         labels = c("none", "infected"))
anno$status <- substr(anno$individual, 1, 1)
anno$status <- factor(anno$status, levels = c("L", "P", "T"),
                      labels = c("latent", "pulmonary", "meningeal"))
colnames(x) <- paste(anno$individual, anno$treatment, sep = ".")
table(anno$treatment, anno$status)
```

## limma analysis (probe-level)

Following limma manual sections 9.3, 9.5.2, and 9.7.

```{r limma-probe}
f <- paste(anno$status, anno$treatment, sep = ".")
f <- factor(f)
design <- model.matrix(~0 + f)
colnames(design) <- levels(f)
colSums(design)
corfit <- duplicateCorrelation(x, design, block = anno$individual)
corfit$consensus
fit <- lmFit(x, design, block = anno$individual, correlation = corfit$consensus)
contrast_matrix <- makeContrasts(
  latent = latent.infected - latent.none,
  pulmonary = pulmonary.infected - pulmonary.none,
  # meningeal = meningeal.infected - meningeal.none,
  interaction = (pulmonary.infected - pulmonary.none) -
                (latent.infected - latent.none),
  levels = design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
results <- decideTests(fit2)
summary(results)
vennDiagram(results)
```

## Mapping probes to genes

Download probe annotation.

```{r probe2id}
ensembl <- useMart(host = "dec2015.archive.ensembl.org",
                   biomart = "ENSEMBL_MART_ENSEMBL",
                   dataset = "hsapiens_gene_ensembl")
probe2id <- getBM(attributes = c("affy_hg_u133_plus_2", "ensembl_gene_id",
                                 "chromosome_name", "status",
                                 "external_gene_name", "gene_biotype"),
                   filters = "affy_hg_u133_plus_2",
                   values = as.character(fData(gse)$ID),
                   mart = ensembl)
head(probe2id)
```

Filtering by chromosome and biotype.

```{r probe-filter-chr-biotype}
probes <- probe2id[probe2id$chromosome_name %in% c(1:22, "X", "Y", "MT") &
                   probe2id$gene_biotype == "protein_coding", ]
nrow(probes)
length(unique(probes$affy_hg_u133_plus_2))
```

Keeping only those probes that unambiguously tag one protein-coding gene.

```{r filter-ambiguous-probes}
genes_per_probe <- table(probes$affy_hg_u133_plus_2)
ambiguous_probes <- names(genes_per_probe)[genes_per_probe > 1]
good_probes <- names(genes_per_probe)[genes_per_probe == 1]
probes_1pergene <- probes[probes$affy_hg_u133_plus_2 %in% good_probes, ]
stopifnot(length(probes_1pergene$affy_hg_u133_plus_2) ==
          length(unique(probes_1pergene$affy_hg_u133_plus_2)))
probes_per_gene <- table(probes_1pergene$ensembl_gene_id)
table(probes_per_gene)
```

Many genes are tagged by more than one probe.
I'll average them to get one value per gene.

```{r average-probes}
x_filtered <- x[probes_1pergene$affy_hg_u133_plus_2, ]
x_filtered <- as.data.frame(x_filtered)
x_filtered$ensembl <- probes_1pergene$ensembl_gene_id
x_mean <- x_filtered %>%
  group_by(ensembl) %>%
  summarise_each(funs(mean)) %>%
  as.data.frame
stopifnot(length(x_mean$ensembl) == length(unique(x_mean$ensembl)))
rownames(x_mean) <- x_mean$ensembl
x_mean <- x_mean %>% select(-ensembl)
```

## limma analysis (gene-level)

Following limma manual sections 9.3, 9.5.2, and 9.7.

```{r limma-gene}
# design matrix defined above in probe-level analysis
colSums(design)
corfit_gene <- duplicateCorrelation(x_mean, design, block = anno$individual)
corfit_gene$consensus
fit_gene <- lmFit(x_mean, design, block = anno$individual,
                  correlation = corfit_gene$consensus)
# contrast matrix defined above in probe-level analysis
contrast_matrix
fit2_gene <- contrasts.fit(fit_gene, contrast_matrix)
fit2_gene <- eBayes(fit2_gene)
results_gene <- decideTests(fit2_gene)
summary(results_gene)
vennDiagram(results_gene)
```

## Exploration

Perform PCA

```{r pca}
pca <- prcomp(t(x_mean), retx = TRUE, center = TRUE, scale. = TRUE)
plot(pca)
```

PC1 is driven by the outlier LTB2.
PC2 is the treatment effect.

```{r pc1vpc2}
plot(pca$x[, "PC1"], pca$x[, "PC2"], col = anno$treatment,
     xlab = "PC1", ylab = "PC2")
text(pca$x[, "PC1"], pca$x[, "PC2"], labels = anno$individual,
     pos = 1, cex = 0.75)
```

Maybe PC3 is starting to get at susceptibility status?

```{r pc2vpc3}
plot(pca$x[, "PC2"], pca$x[, "PC3"], col = anno$treatment,
     xlab = "PC2", ylab = "PC3")
text(pca$x[, "PC2"], pca$x[, "PC3"], labels = anno$individual,
     pos = 1, cex = 0.75)
```

```{r pc3-status}
boxplot(pca$x[, "PC3"] ~ anno$status, ylab = "PC3")
points(pca$x[, "PC3"] ~ anno$status, col = anno$treatment)
```

How does hierarchical clustering sort the outlier?

```{r hclust-cor}
heatmap.2(cor(x_mean), trace = "none", margins = c(7, 7))
```

LTB2.none does not cluster with either the infected or non-infected samples.
They also note in their paper that LTB2 is an outlier, however they only visualize the fold change (Fig. S1).

How different is this sample? Looking at the top 2000 most variable genes.

```{r hclust-variable}
select_var <- function(x, y) {x[order(apply(x, 1, var), decreasing = TRUE), ][1:y, ]}
x_mean_sub <- select_var(x_mean, 2000)
heatmap.2(as.matrix(x_mean_sub), trace = "none", margins = c(7, 7))
```

I think LTB2.none needs to be excluded entirely.

## Session information

```{r info}
sessionInfo()
```
