---
title: "Analysis of Thuong et al., 2008"
date: 2016-05-27
output:
  html_document:
    toc: true
    toc_float: true
---

**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options}
library("knitr")
opts_chunk$set(cache = FALSE)
```

http://www.ncbi.nlm.nih.gov/pubmed/19057661

http://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1000229#s2

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE11199

http://www.bioconductor.org/packages/release/bioc/html/affy.html

http://bioconductor.org/packages/release/bioc/vignettes/GEOquery/inst/doc/GEOquery.html

http://www.bioconductor.org/packages/release/data/annotation/html/hgu133plus2.db.html

## Setup

```{r packages, message=FALSE, warning=FALSE}
library("GEOquery")
library("affy")
library("hgu133plus2.db")
library("limma")
library("tidyr")
```

## Analyze starting from processed data


```{r download}
geo_fname <- "../data/GSE11199.rds"
if (!file.exists(geo_fname)) {
  gse <- getGEO(GEO = "GSE11199")
#   class(gse)
#   length(gse)
  gse <- gse[[1]]
  saveRDS(gse, geo_fname)
} else {
  gse <- readRDS(geo_fname)
}
class(gse)
pData(gse)[1, ]
```

```{r}
x <- exprs(gse)
anno <- pData(gse)[, 1]
anno <- as.data.frame(anno)
anno <- separate(anno, col = anno, into = c("individual", "treatment"))
anno$treatment <- factor(anno$treatment, levels = c("unstim", "stim"),
                         labels = c("none", "infected"))
anno$status <- substr(anno$individual, 1, 1)
anno$status <- factor(anno$status, levels = c("L", "P", "T"),
                      labels = c("latent", "pulmonary", "meningeal"))
table(anno$treatment, anno$status)
```

Convert probe IDs to Ensembl gene IDs.

```{r}
id_table <- as.data.frame(hgu133plus2ENSEMBL)
dim(id_table)
length(unique(id_table$probe_id))
length(unique(id_table$ensembl_id))
sum(id_table$probe_id %in% rownames(x))
```

```{r}
probe_mappings <- select(hgu133plus2.db, keys = rownames(x),
                         columns = c("ENSEMBL", "SYMBOL"))
```


Following limma manual sections 9.3, 9.5.2, and 9.7.

```{r}
f <- paste(anno$status, anno$treatment, sep = ".")
f <- factor(f)
design <- model.matrix(~0 + f)
colnames(design) <- levels(f)
colSums(design)
corfit <- duplicateCorrelation(x, design, block = anno$individual)
corfit$consensus
fit <- lmFit(x, design, block = anno$individual, correlation = corfit$consensus)
contrast_matrix <- makeContrasts(
  latent = latent.infected - latent.none,
  pulmonary = pulmonary.infected - pulmonary.none,
  # meningeal = meningeal.infected - meningeal.none,
  interaction = (pulmonary.infected - pulmonary.none) -
                (latent.infected - latent.none),
  levels = design)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)
results <- decideTests(fit2)
summary(results)
vennDiagram(results)
```

## Analyze starting from raw data

Download.

```{r}
getGEOSuppFiles(GEO = "GSE11199")
untar("GSE11199/GSE11199_RAW.tar", exdir = "GSE11199")
```

Import.

```{r}
raw <- ReadAffy(filenames = Sys.glob("GSE11199/*CEL.gz"), compress = TRUE)
```

Pre-process.
The only information they provided on GEO was "Raw CEL intensity data were RMA normalized using R/Bioconductor".
Thus I just used the recommendation from the affy package example for the other steps.

```{r}
if (file.exists("thuong2008.rds")) {
  eset <- readRDS("../data/thuong2008.rds")
} else {
  eset <- expresso(raw,
                   bgcorrect.method = "rma",
                   normalize.method = "quantiles",
                   pmcorrect.method = "pmonly",
                   summary.method = "medianpolish")
  saveRDS(eset, "../data/thuong2008.rds")
}
```

## Compare the two approaches

My pre-processing obtains results very similar to theirs.

```{r}
dim(x)
y <- exprs(eset)
x[1:5, 1:5]
y[1:5, 1:5]
plot(x[, 1], y[, 1])
cor(x[, 1], y[, 1])
z <- cor(cbind(x, y))
dim(z)
min(z)
median(z)
```

## Session information

```{r info}
sessionInfo()
```
